## Preliminaries

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
library(tmap)
library(ggplot2)
```

## Traffic Analysis Zones

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp")
```

Project object:
```{r}
ggh_taz <- ggh_taz %>%
  st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
```

Extract the IDs for all TAZ:
```{r}
ghta_taz_id <- ggh_taz$GTA06
```

## Planning boundaries

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp")
```

Project object:
```{r}
ggh_pd <- ggh_pd %>%
  st_transform(crs = 32617)
```

Adding region names to the object as they are not included in the origin data:
(http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf):
```{r}
ggh_pd <- ggh_pd %>%
  mutate(REGION_name = 
           ifelse(REGION == 1, "Toronto", 
                  ifelse(REGION == 2, "Durham", 
                         ifelse(REGION == 3, "York", 
                                ifelse(REGION == 4, "Peel", 
                                       ifelse(REGION == 5, "Halton", 
                                              ifelse(REGION == 6, "Hamilton", 
                                                     ifelse(REGION == 11, "Niagara", 
                                                            ifelse(REGION == 12, "Waterloo", 
                                                                   ifelse(REGION == 13, "Guelph", 
                                                                          ifelse(REGION == 14, "Wellington", 
                                                                                 ifelse(REGION == 15, "Orangeville", 
                                                                                        ifelse(REGION == 16, "Barrie", 
                                                                                               ifelse(REGION == 17, "Simcoe", 
                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
                                                                                                             ifelse(REGION == 19, "Peterborough City", 
                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
                                                                                                                           ifelse(REGION == 21, "Orillia", 
                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
                                                                                                                                         ifelse(REGION == 23, "Brantford", 
                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
```

add 'REGION' from the planning distract object to the `ggh_taz`. 
```{r}
#buffering to clean up the edges of these objects -> there wre issues using st_join. 
ggh_taz <- st_buffer(ggh_taz, dist=0)
ggh_pd <- st_buffer(ggh_pd, dist=0)
```


```{r}
ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
```

## Origin destination matrices for place of residence and school destination (by person)
Read exported TTS 2011 and 2016 tables. The query was a cross-tabulation of <!--"persons"--> "trips" by 2006 zone of residence and 2006 zone of school using motorized modes ("mode_prime In B, D, G, J, M, O, P, S, T, U, 9" -- this is only one of four options, school bus, auto care passenger, taxi passenger, or transit) or non-motorized ("mode_prime In C, W" -- cycling or walking). The data was filtered to only include Hamilton residential households, "Trip purpose of destination" as school, the student status as full/part-time student, and age In 5-14:

```{r}
od_trips_2011_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
od_trips_2016_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2011_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2016_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
# od_trips_2011 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2011.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)
# od_trips_2016 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2016.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)

```

Find the position in the table where the information on work trips begins:
```{r}
# idx <- which(od_trips_2011$X1 == "TABLE    : stu_stat (Full time)")
idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")
```

Slice the table to obtain the work trips (remove heading information):
```{r}
od_trips_2011_motor <- od_trips_2011_motor %>% 
  slice((idx+2):n())
```

Separate the zone identifiers and the trips, and convert to numeric:
```{r}
od_trips_2011_motor <- od_trips_2011_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

```{r}
idx <- which(od_trips_2011_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>% 
  slice((idx+2):n())

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

Repeat the same steps for 2016:
```{r}
idx <- which(od_trips_2016_motor$X1 == "COLUMN : gta06_sch")

od_trips_2016_motor <- od_trips_2016_motor %>% 
  slice((idx+2):n())

od_trips_2016_motor <- od_trips_2016_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))

idx <- which(od_trips_2016_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  slice((idx+2):n())

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

The resulting objects `od_trips_2011`.. and  `od_trips_2016`.. has the number of trips made by students aged 5-14 who are full-time/part-time students by place of residence (Origin) and place of school (Destination) using the GTA06 zoning system. 

Now let's see how many students travel outside of their catchment. 
```{r}
#load catchments
load(file="data-products/ALL_CATCHS.RData")
CATCH_2011_Public <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
CATCH_2011_Cath <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Public <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Cath <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))

#assign a catchment id (from elem, 2011) to od_trips_2011 taz ids, what I'm doing here is creating a column of lists -- the lists contain all the GTA06 within a single catchment. I do this for all 4 catchments (2011, 2016 and pub, cath). GTA06s that are within  multiple catchments are repeated i.e., a GTA06 that is in 3 catchments spatially is withinin 3 lists. 
ham_taz_catch_2011_public <-st_join(ham_taz,CATCH_2011_Public, join=st_intersects) %>% select(c("GTA06", "CID")) %>% st_drop_geometry() %>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2011_cath <-st_join(ham_taz,CATCH_2011_Cath, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2016_public <-st_join(ham_taz,CATCH_2016_Public, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2016_cath <-st_join(ham_taz,CATCH_2016_Cath, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
```

```{r}
## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023

# Function to check if values in a row of df1 are within any list in df2
check_within_lists <- function(df1, df2) {
  # Initialize an empty list to store the results
  result_list <- list()

  # Check if values in a row of df1 are within any list in df2
  for (i in 1:nrow(df1)) {
    matching_ids <- which(sapply(df2$check_list, function(lst) all(df1[i, ] %in% unlist(lst))))
    if (length(matching_ids) > 0) {
      result_list[[i]] <- data.frame(result = TRUE, df1[i, ], CID = matching_ids)
    } else {
      result_list[[i]] <- data.frame(result = FALSE, df1[i, ], CID = NA)
    }
  }

  # Combine the results into a single dataframe
  result_df <- do.call(rbind, result_list)

  return(result_df)
}

# # Example usage is commented out:
# # Assuming you have two data frames named 'df1' and 'df2'
# # df1 has columns 'column1' and 'column2' with numeric values
# # df2 has columns 'id' and 'check_list' with lists of numeric values created using I() function
# df1 <- data.frame(column1 = c(1, 2, 3, 4, 5), column2 = c(10, 20, 30, 40, 50))
# df2 <- data.frame(id = c(1, 2, 3), check_list = I(list(c(1, 10), c(3, 30), c(1, 6, 10, 4, 40))))
# 
# # Call the function
# check_within_lists(df1, df2)
```
```{r, warning=FALSE}
## FOR 2011
#the origin-destination trips made within hamilton to hamilton schools (all in TTS Region `6`).
ODs <- data.frame(Origin = od_trips_2011_motor$Origin,
                 Destination = od_trips_2011_motor$Destination)
#the list of GTA06 zones within each "CID" id)
LISTS <- data.frame(CID = ham_taz_catch_2011_public$CID,
                  check_list = I(ham_taz_catch_2011_public$GTA06_in_catch))

LISTS_catch_2011_public <- check_within_lists(df1 = ODs, df2= LISTS)

LISTS <- data.frame(CID = ham_taz_catch_2011_cath$CID,
                  check_list = I(ham_taz_catch_2011_cath$GTA06_in_catch))
LISTS_catch_2011_cath <- check_within_lists(df1 = ODs, df2= LISTS)

## FOR 2016
ODs <- data.frame(Origin = od_trips_2016_motor$Origin,
                 Destination = od_trips_2016_motor$Destination)

LISTS <- data.frame(CID = ham_taz_catch_2016_public$CID,
                  check_list = I(ham_taz_catch_2016_public$GTA06_in_catch))

LISTS_catch_2016_public <- check_within_lists(df1 = ODs, df2= LISTS)

LISTS <- data.frame(CID = ham_taz_catch_2016_cath$CID,
                  check_list = I(ham_taz_catch_2016_cath$GTA06_in_catch))
LISTS_catch_2016_cath <- check_within_lists(df1 = ODs, df2= LISTS)

rm(ODs,LISTS)
```

```{r}
LISTS_catch_2011_public <- LISTS_catch_2011_public %>% 
  merge(od_trips_2011_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2011_cath <- LISTS_catch_2011_cath %>% 
  merge(od_trips_2011_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2016_public <- LISTS_catch_2016_public %>% 
  merge(od_trips_2016_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2016_cath <- LISTS_catch_2016_cath %>% 
  merge(od_trips_2016_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))
```
```{r}
(LISTS_catch_2011_public$Trips_perc*LISTS_catch_2011_public$result) %>% sum()
(LISTS_catch_2011_cath$Trips_perc*LISTS_catch_2011_cath$result) %>% sum()
(LISTS_catch_2016_public$Trips_perc*LISTS_catch_2016_public$result) %>% sum()
(LISTS_catch_2016_cath$Trips_perc*LISTS_catch_2016_cath$result) %>% sum()

# these sums make sense, almost all CIDs (minus 1 small CID in the public catchments) are represented... e.g., checking the count of 'LISTS_catch_2011_public$CID %>% unique() %>% count()' against how many invidual CIDs there are. 
```
Okay, so here we see approximately ~80% of trips that are 'motorized' made by students in Hamilton (5-14 age) to hamilton schools are made within catchment. 

Interestingly, if we repeat this process for the non-motorized trips -- almost 100% of trips are made within catchment! This is an interesting but intuitive finding. Using a wider-range modes allows you to reach larger distance. In a way, those who walk/cycle have a reduced luxury of school-seat choice. 

## Calculating car travel time for OD of hamilton TAZs

Libraries:
```{r setup, include=FALSE, eval=FALSE}
library(disk.frame)
library(ggplot2)
library(gridExtra)
library(httr)
library(knitr)
#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
library(kableExtra)
#install.packages("devtools")
#install.packages("Rtools")
#devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
library(lodes)
library(progress)
library(purrr)
library(r5r)
#library(stplanr)
library(tidyr)
library(tmap)
library(zoo) # for rollapplyr

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx512m")
options(future.globals.maxSize = Inf)
```

Format the destination and origin points (ID, Long, Lat):
```{r, eval=FALSE}
ham_taz_coords <-  
  cbind(ham_taz %>% st_drop_geometry()%>% transmute(GTA06), st_centroid(ham_taz) %>% st_transform(crs = 4326) %>% st_coordinates()) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") 
```
Set-up directories and download the osm network:
```{r set up r5 path, include=FALSE, eval=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-inputs/TTS/r5_graph")
r5_path <- file.path("data-inputs/TTS/r5_graph")
```

```{r download data, include=FALSE, eval=FALSE}
# downloading ontario osm in the correct format
# download.file(url = paste0("https://download.geofabrik.de/north-america/canada/ontario-latest.osm.pbf"),
#               destfile = file.path(r5_path, "osm.pbf"),
#               mode = "wb")

```
Set Up R5 Routing:
```{r build graph, include = FALSE, eval=FALSE}
# NOTE: I had signficant issues getting this function to run. Since the most recent r5r update (before Nov ~25 2022 when I was working on this), I believe the setup_r5 requires smaller road networks. So putting the full ontario geofabrik extract no longer works for me. I used http://share.mapbbcode.org/ to export the Hamilton area, and then OSMCONVERT to convert the .osm file into a .pbf.
r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)

```

Now calculate the OD matrix, for car travel time from all origins (residential GTA06) to all destinations (school GTA06):
```{r, eval=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(ham_taz_coords)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = ham_taz_coords,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2022-11-16 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 120)
  
  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car <- as.data.frame(output_df)

save("output_OD_car", file = "data-inputs/TTS/output_OD_car.Rdata")
#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
```

```{r, eval=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(ham_taz_coords)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = ham_taz_coords,
                          mode = c("WALK"), 
                          departure_datetime = as.POSIXct(strptime("2022-11-16 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 120)
  
  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk <- as.data.frame(output_df)

save("output_OD_walk", file = "data-inputs/TTS/output_OD_walk.Rdata")
```
Java garbage collector:
```{r, eval=FALSE}
stop_r5(r5_HAM)
rJava::.jgc(R.gc = TRUE)
```

## Inspecting the travel times and calculating TLD
```{r}
load(file = "data-inputs/TTS/output_OD_car.Rdata")
summary(output_OD_car)
#we'll set 0 travel times to equal 7 min for inter-zonal travel, half of the 1st Qu. travel time.

load(file = "data-inputs/TTS/output_OD_walk.Rdata")
summary(output_OD_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
```

Trip length distribution (cutting the travel_time into 150 intervals and summing all trips within each interval). The TLD is a probability density function for likelihood of travel informed by travel time:
```{r}
#join travel_time to 2011 and 2016 OD trips
od_trips_2011_motor <- od_trips_2011_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06") 

od_trips_2011_motor <- od_trips_2011_motor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (7)*AREA/max(od_trips_2011_motor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) #~7 mins is Q1 for 'tl_2016_motor.

od_trips_2016_motor <- od_trips_2016_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06") 

od_trips_2016_motor <- od_trips_2016_motor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (7)*AREA/max(od_trips_2016_motor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) ##~7 mins is Q1 for 'tl_2016_motor.

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06") 

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (15)*AREA/max(od_trips_2011_nonmotor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) ##~15 mins is Q1 for 'tld_2016_nonmotor.

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06") 

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (15)*AREA/max(od_trips_2016_nonmotor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) #~15 mins is Q1 for 'tld_2016_nonmotor.
```

```{r}
tld_2011_motor <- od_trips_2011_motor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_motor <- od_trips_2016_motor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2011_nonmotor <- od_trips_2011_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_nonmotor <- od_trips_2016_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))
```

Plot the TLD, motorized:
```{r}
ggplot(data = tld_2011_motor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

ggplot(data = tld_2016_motor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

summary(tld_2011_motor)
summary(tld_2016_motor)
```

Plot the TLD, nonmotorized:
```{r}
ggplot(data = tld_2011_nonmotor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

ggplot(data = tld_2016_nonmotor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

summary(tld_2011_nonmotor)
summary(tld_2016_nonmotor)
```


Lets fit a curve.

First, let's transform the data so each travel time for each OD trip is a row
```{r data-for-Hamilton-trips, include=FALSE}
#for 2011
all_tt_2011_motor <- od_trips_2011_motor  %>% 
  dplyr::select(Trips, travel_time)
sum((od_trips_2011_motor$Trips))
all_tt_2011_motor <- all_tt_2011_motor[rep(seq_len(dim(all_tt_2011_motor)[1]), all_tt_2011_motor$Trips), 2]

all_tt_2011_nonmotor <- od_trips_2011_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2011_nonmotor$Trips))
all_tt_2011_nonmotor <- all_tt_2011_nonmotor[rep(seq_len(dim(all_tt_2011_nonmotor)[1]), all_tt_2011_nonmotor$Trips), 2]

#for 2016
all_tt_2016_motor <- od_trips_2016_motor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_motor$Trips))
all_tt_2016_motor <- all_tt_2016_motor[rep(seq_len(dim(all_tt_2016_motor)[1]), all_tt_2016_motor$Trips), 2]

all_tt_2016_nonmotor <- od_trips_2016_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_nonmotor$Trips))
all_tt_2016_nonmotor <- all_tt_2016_nonmotor[rep(seq_len(dim(all_tt_2016_nonmotor)[1]), all_tt_2016_nonmotor$Trips), 2]

```

```{r fitting-impedance-function-2011-2016-1, cache=TRUE}
library(fitdistrplus)
descdist(all_tt_2011_motor, discrete=FALSE)
descdist(all_tt_2016_motor, discrete=FALSE)

descdist(all_tt_2011_nonmotor, discrete=FALSE)
descdist(all_tt_2016_nonmotor, discrete=FALSE)
```

```{r fitting-impedance-function-2011-2016-2, cache=TRUE}
gamma2011_motor <- fitdistrplus::fitdist(data=all_tt_2011_motor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2011_motor <- fitdistrplus::fitdist(data=all_tt_2011_motor, "exp", method="mme", optim.method="Brent")

gamma2016_motor <- fitdistrplus::fitdist(data=all_tt_2016_motor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2016_motor <- fitdistrplus::fitdist(data=all_tt_2016_motor, "exp", method="mme", optim.method="Brent")

plot(gamma2011_motor)
plot(exp2011_motor)

plot(gamma2016_motor)
plot(exp2016_motor)

gamma2011_motor$aic #gamma has the lowest aic and bic, thus we pick gamma distribution. 
exp2011_motor$aic

gamma2016_motor$aic
exp2016_motor$aic
```

```{r fitting-impedance-function-2011-2016-2, cache=TRUE}
gamma2011_nonmotor <- fitdistrplus::fitdist(data=all_tt_2011_nonmotor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2011_nonmotor <- fitdistrplus::fitdist(data=all_tt_2011_nonmotor, "exp", method="mme", optim.method="Brent")

gamma2016_nonmotor <- fitdistrplus::fitdist(data=all_tt_2016_nonmotor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2016_nonmotor <- fitdistrplus::fitdist(data=all_tt_2016_nonmotor, "exp", method="mme", optim.method="Brent")

plot(gamma2011_nonmotor)
plot(exp2011_nonmotor)

plot(gamma2016_nonmotor)
plot(exp2016_nonmotor)

gamma2011_nonmotor$aic #gamma and exp have similar aic and bic. empirically exp is often selected, so we pick it here too.
exp2011_nonmotor$aic

gamma2016_nonmotor$aic
exp2016_nonmotor$aic
```
```{r save-impedance-plot, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2011-motor.png")
plot(gamma2011_motor)
dev.off()
```
```{r save-impedance-plot-2, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2016-motor.png")
plot(gamma2016_motor)
dev.off()
```
```{r save-impedance-plot-3, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2011-nonmotor.png")
plot(exp2011_nonmotor)
dev.off()
```
```{r save-impedance-plot-4, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2016-nonmotor.png")
plot(exp2016_nonmotor)
dev.off()
```

And save the gamma functions, to use in the analysis
```{r}
save(gamma2011_motor, file="data-inputs/TTS/gamma2011_motor.Rdata")
save(gamma2016_motor, file="data-inputs/TTS/gamma2016_motor.Rdata")
save(exp2011_nonmotor, file="data-inputs/TTS/exp2011_nonmotor.Rdata")
save(exp2016_nonmotor, file="data-inputs/TTS/exp2016_nonmotor.Rdata")
```

